\chapter*{Chapter 3}
\label{sec:chapter_3}

\latline
  {Itaque ibi coeperunt Traiani munire locum, quem Aeneas ex}
  { And indeed there the Trojans began to found a place, which Aeneas }
  {150}
  { CorrectedTranslation }
  { \begin{enumerate}
  	\item mun\={\i}re:  fortify, stringthen found
  \end{enumerate} }


\latline
  {nomine uxoris suae, Latini regis filiae, quae iam ante desponsa Turno}
  { called after the name of his own wife, the daughter of the king of the Latins who prior to this had been betrothed to Turnus, }
  {150}
  { CorrectedTranslation }
  { \begin{enumerate}
  	\item 
  \end{enumerate} }


\latline
  {Lavinium nominavit.  At vero Amata, Latini regis uxor, cum}
  { Lavinium.  But indeed, Amata, the wife of King Latinus, since}
  {150}
  { CorrectedTranslation }
  { Notes }


\latline
  {indigne ferret Laviniam, repudiato Turno consobrino suo, Troiano}
  { she was believing it unfitting that Lavinia, having repudiated her own cousin Turnus,  }
  {150}
  { CorrectedTranslation }
  { \begin{enumerate}
  	\item  consobrinus, a, um, c:  cousin on mother's side
  	\item ferret = ferre +t = subjunctive imperfect + cum to make it a causal subjunctive
  \end{enumerate} }


\latline
  {[\textbf{5}] advenae collocatam esse, Turnum ad arma concitavit.  Isque mox,}
  { was being promised to the Trojan stranger, agitated Turnus to arms.  And soon he, }
  {150}
  { CorrectedTranslation }
  { \begin{enumerate}
  	\item advena, -ae, c:  foreigner, stranger
  	\item collocatus:  placed, stationed
  	\item concito(1):  to excite, stir up
  \end{enumerate} }


\latline
  {coacto Rutulorum exercitu, tetendit in agrum Laurentem.  Et Latinus,}
  { having led the Rutulian army, went toward the Laurentian country.  But Latinus, }
  {150}
  { CorrectedTranslation }
  { \begin{enumerate}
  	\item ad (city) means ``in the region of (city)''; city in the accusative means to the city
  \end{enumerate} }


\latline
  {adversus eum pariter cum Aenea progressus inter proeliantes, circumventus}
  { against him, had marched side-by-side with Aeneas among the battling men, had been surrounded }
  {150}
  { CorrectedTranslation }
  { \begin{enumerate}
  	\item adversus:  against, contrary to
  	\item pariter:  equally, totgether
  	\item progredior, progredi, progressus:  to march, set out
  	\item circumvenio, circumven\={\i}re, circumveni, circumventus:  to encircle, assail, attack
  \end{enumerate} }


\latline
  {est occisusque.  Nec tamen, amisso socero, Aeneas Rutulis}
  { and had fallen in battle.  Nevertheless, with his comerade having been lost, Aeneas  }
  {150}
  { CorrectedTranslation }
  { \begin{enumerate}
  	\item occido, occidere, occidi, occasus:  to set, to die, to perish
  \end{enumerate} }


\latline
  {obsistere destitit, tandemque Turnum interemit.  Hostibus fusis}
  { did not cease standing against the Rutulians and halted Turnus.  With the armies confused }
  {150}
  { CorrectedTranslation }
  { \begin{enumerate}
  	\item obsist\={o}, obsistere, obstit\={i}, obstitus:  to resist
  	\item desisto, desistere, destiti, destitus:  to stop
  	\item intermitto, intermittere, intermisi, intermissus:  interrupt, omit, stop
  	\item fusus, -a, -um:  spread out, flowing 
  \end{enumerate} }


\latline
  {[\textbf{10}] fugatisque, victor Lavinium se cum suis recepit, consensuque omnium}
  { and routed, the victor withdrew himself and his armies to Lavinium and by consent of all 	 }
  {150}
  { CorrectedTranslation }
  { \begin{enumerate}
  	\item fugatus, -a, -um: put to flight
  	\item consensus, us, m: agreement, opinion
  \end{enumerate} }


\latline
  {Latinorum rex declaratus est.}
  { the Latins he was declared the king. }
  {150}
  { CorrectedTranslation }
  { Notes }


